---
title: "Plots From Large Datasets"
output: html_notebook
---


## Disclaimer
Make a few experiments with plots with large datasets


## Devices
Different devices such as `png`, `jpeg` and others require `xorg-dev` library. 


## Experiments

We generate a reasonably sized dataset and try different potting options

```{r}
n_nr_obs <- 10^5
n_inter <- 3.24
n_slope <- 0.7
x <- runif(n_nr_obs)
y <- n_inter + n_slope * x + rnorm(n_nr_obs)
```


## Plots
Normal inline plotting looks as follows

```{r}
plot(x,y)
```

## Plots to Files
Different devices generate different file sizes. 

```{r}
s_file_stem <- "Rplot"
s_file_dev <- "png"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
png(filename = s_out_path)
plot(x,y)
dev.off()
```

The generated file has the following properties

```{r}
(df_png_finfo <- file.info(s_out_path))
```

This file was generated as a benchmark for the filesize which is: `r df_png_finfo$size`

## Different Devices

### pdf

```{r}
s_file_stem <- "Rplot"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path)
plot(x,y)
dev.off()
```

The file size here is

```{r}
(df_pdf_finfo <- file.info(s_out_path))
```



### Postscript

```{r}
s_file_stem <- "Rplot"
s_file_dev <- "ps"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
postscript(file = s_out_path)
plot(x,y)
dev.off()
```

The file size here is

```{r}
(df_pdf_finfo <- file.info(s_out_path))
```


## Available Options
Because the filesize of `pdf()` was smaller, we first look at options for this device. Options can be listed by

```{r}
pdf.options()
```

* width and height

```{r}
s_file_stem <- "Rplothw14"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, width = 14, height = 14)
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))
```

Changing the height and width does not make the file size smaller

```{r}
s_file_stem <- "Rplotps2"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, pointsize = 2)
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))

```

```{r}
s_file_stem <- "Rplotps24"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, pointsize = 24)
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))
```

```{r}
s_file_stem <- "Rplotcmg"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, colormodel = "grey")
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))
```


## More tools

### compactPDF
The function `tools::compactPDF()` did not work


### 
From https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Murrell.pdf


