---
title: "Plots From Large Datasets"
output: html_notebook
---

```{r}
### # define vector of required packages
vec_req_pkgs <- c("tibble", "ggplot2")
### # extract vector of all installed packages
vec_inst_pkgs <- installed.packages()
### # install required packages, if necessary
for (p in vec_req_pkgs){
  if (!p %in% vec_inst_pkgs)
    install.packages(pkgs = p, repos = "https://cran.rstudio.com")
}

```


## Disclaimer
Make a few experiments with plots with large datasets


## Devices
Different devices such as `png`, `jpeg` and others require `xorg-dev` library. 


## Experiments

We generate a reasonably sized dataset and try different potting options

```{r}
n_nr_obs <- 10^5
n_inter <- 3.24
n_slope <- 0.7
x <- runif(n_nr_obs)
y <- n_inter + n_slope * x + rnorm(n_nr_obs)
tbl_plot_data <- tibble::tibble(x = x, y = y)
```


## Plots
Normal inline plotting looks as follows

```{r}
plot(x,y)
```

## Plots to Files
Different devices generate different file sizes. 

```{r}
s_file_stem <- "Rplot"
s_file_dev <- "png"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
png(filename = s_out_path)
plot(x,y)
dev.off()
```

The generated file has the following properties

```{r}
(df_png_finfo <- file.info(s_out_path))
```

This file was generated as a benchmark for the filesize which is: `r df_png_finfo$size`

## Different Devices

### pdf

```{r}
s_file_stem <- "Rplot"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path)
plot(x,y)
dev.off()
```

The file size here is

```{r}
(df_pdf_finfo <- file.info(s_out_path))
```



### Postscript

```{r}
s_file_stem <- "Rplot"
s_file_dev <- "ps"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
postscript(file = s_out_path)
plot(x,y)
dev.off()
```

The file size here is

```{r}
(df_pdf_finfo <- file.info(s_out_path))
```


## Available Options
Because the filesize of `pdf()` was smaller, we first look at options for this device. Options can be listed by

```{r}
pdf.options()
```

* width and height

```{r}
s_file_stem <- "Rplothw14"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, width = 14, height = 14)
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))
```

Changing the height and width does not make the file size smaller

```{r}
s_file_stem <- "Rplotps2"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, pointsize = 2)
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))

```

```{r}
s_file_stem <- "Rplotps24"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, pointsize = 24)
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))
```

```{r}
s_file_stem <- "Rplotcmg"
s_file_dev <- "pdf"
s_out_path <- paste(s_file_stem, s_file_dev, sep = ".")
pdf(file = s_out_path, colormodel = "grey")
plot(x,y)
dev.off()
(df_pdf_finfo <- file.info(s_out_path))
```


## More tools

### compactPDF
The function `tools::compactPDF()` did not work


### grid.raster
From https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Murrell.pdf, we get the following example code-junk

```{r}
library(grid)
z <- matrix(runif(500*500), ncol=500)
pdf("image.pdf")
image(z, col=grey(0:99/100))
dev.off()
pdf("gridraster.pdf")
grid.raster(z, interp=FALSE)
dev.off()
file.info("image.pdf", "gridraster.pdf")["size"]
```


To use the `grid.raster()` function, we have to convert the scatter-plot into a raster matrix. 


### smoothscatter
Avoiding effects of overplotting with a smoothed version of the scatterplot. See https://www.stat.ubc.ca/~jenny/STAT545A/block91_latticeGraphics.html for more details.

```{r}
s_out_path <- "RsmoothScatter.pdf"
pdf(s_out_path)
smoothScatter(y ~ x, colramp = colorRampPalette(c("white", blues9)), xlim = c(0,1), ylim = c(0, 8))
dev.off()
file.info(s_out_path)["size"]
```



### ggsave
The plot is drawn with `ggplot2` and `ggsave()` is used to save the plot to a file

```{r}
s_out_path <- "Rggplot2.pdf"
library(ggplot2)
ggplot(tbl_plot_data, aes(x,y)) + geom_point()
ggsave(s_out_path, width = 4, height = 4, dpi = "screen")
(df_ggplot_pdf_finfo <- file.info(s_out_path))
```

Using a different file extension

```{r}
s_out_path <- "Rggplot2.png"
library(ggplot2)
ggplot(tbl_plot_data, aes(x,y)) + geom_point()
ggsave(s_out_path, width = 4, height = 4, dpi = "screen")
(df_ggplot_png_finfo <- file.info(s_out_path))
```

More options of `ggsave()`

```{r}
?ggsave
```

Based on the body of function `ggsave()`, the plot devices are generated by `ggplot2:::plot_dev()` and inside that function `grDevices::png()` is called to produce a `png`-plot. From the help of `grDevices::png()` it is clear that this function also depends on `X11` functionality. 


### CairoPng
From Chang2013 (R Graphics Cookbook) p. 345 ...

```{r}
s_out_path <- "RplotCairo.png"
Cairo::CairoPNG(filename = s_out_path)
plot(x, y)
dev.off()
(df_cairo_png_info <- file.info(s_out_path))
```

Compilation of `Cairo` on Ubuntu 16.04 lts fails with the following error

```
checking for pkg-config... no
configure: CAIRO_LIBS is unset, attempting to guess it.
configure: CAIRO_CFLAGS=
checking if R was compiled with the RConn patch... no
checking cairo.h usability... no
checking cairo.h presence... no
checking for cairo.h... no
configure: error: Cannot find cairo.h! Please install cairo (http://www.cairographics.org/) and/or set CAIRO_CFLAGS/LIBS correspondingly.
ERROR: configuration failed for package 'Cairo'
* removing '/home/zws/lib/R/library/Cairo'

Die heruntergeladenen Quellpakete sind in
        '/tmp/Rtmpr8wC2N/downloaded_packages'
Warnmeldung:
In install.packages("Cairo", repos = "https://cran.rstudio.com") :
  Installation des Pakets 'Cairo' hatte Exit-Status ungleich 0
```

As can be seen from https://stackoverflow.com/questions/51940792/r-cairo-package-installation-failed-in-ubuntu-18-04, X11 and other libraries are used to compile `Cairo`. 
